The Creation of a Programming Language

### 2022-10-23 paper.tex
add mandelbrot to title-page
add figures to paper

\documentclass[12pt,a4paper]{article}
\usepackage{listings,xcolor,setspace,hyperref,dirtree}
\usepackage{catchfile,pgffor}
\usepackage[margin=2.5cm]{geometry} % Set page margin to 2.5cm
\usepackage{times} % Use Times New Roman font
\usepackage[style=alphabetic,backend=biber]{biblatex}

\addbibresource{bibliography.bib}

@@

% Read contents of files.txt into \files
\CatchFileDef{\files}{files.txt}{\endlinechar=-1}

% Include C code from file #1 from lines #2 to #3
\newcommand{\code}[3] {
    \lstset{language=C,numbers=left,basicstyle=\scriptsize,
        frame=single,title=#1,keywordstyle=\color{blue},stepnumber=5,
        stringstyle=\color{green},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#}}
    \lstinputlisting[firstnumber=#2,firstline=#2,lastline=#3]{#1}
}

@@

\newcommand{\pagelabel}[1]{\phantomsection\label{#1}}

@@

    \vspace*{2.5cm}
    \Huge
    \textbf{The Creation of a Programming Language}

    \Large

    \vspace{1.5cm}
    Matura Paper by\\ \textbf{Florian Donnelly}

@@

@@

% the name of the programming language
\newcommand{\name}{\emph{TestScript }}

@@

A programming language is itself only a description or specification, i.e. a
set of grammar rules defining how single instructions and entire programs
can be written. Source code, which is a program written in a certain language,
has to be understood by the computer hardware in order to run. It can either
be translated to machine-native bytecode with the help of a compiler, or
read line-by-line and interpreted by another program known as interpreter.
Such compilers or interpreters are programs on their own that are classified
as programming language implementation. The focus of this paper will be to
create such a programming language implementation, more specific an interpreter
of the custom-made \name language.

\subsection{Goals}
The focus of this project is learning about the internal design of interpreters
by creating a custom interpreter named \name interpreter, building upon
known strategies and algorithms used in existing programming languages.
The goal of this paper is to further understand the inner workings of
interpreters, stepping through human-readable source code and computing
desired results.

Modern programming languages often times provide additional tools aside from
the implementation. This includes dependency management tools to
easily create projects extending upon existing ones, or a web platforms for
people to share their code, all of which is not subject of this paper.
The \name implementation will also not include many features seen in modern
languages, as those are often created in years of time spent planning and
programming.

@@

        Self-interpreters are closely related to self-hosting compilers. These
        are in turn compilers which are written in the language they compile.
        An example of a compiler written in itself is rustc, the Rust
        compiler\cte{rustc}, which was bootstrapped by the OCaml language.
\end{itemize}
This paper will implement a mixture form of an AST compiler and bytecode
interpreter. This is not practical in terms of improving \name, but serves
the purpose of combining different features of different interpreters into one
and therefore extending the scope of the project.

@@

\subsection{The Art of Abstraction}
When dealing with problems in general, it is helpful to map out and think about
the way of approaching the problem first. A common approach is to distill the
problem situation down to a simpler one and extending the solution to take into
account all important aspects later.

@@

\subsection{A Project Overview}

\subsubsection{File Structure}
Here is a file tree overview of the project which will be referenced
several times in the following sections of this paper. In the depicted file tree
only source files are shown. No build files, executable files or files that
are of no further importance are depicted.
\begin{flushleft}
    \scriptsize\singlespacing\setlength{\DTbaselineskip}{9pt}
    \dirtree{%
        .1 project root.
        .2 .gitignore.
        .2 README.md.
        .2 Makefile.
        .2 mapop.gperf.
        .2 pseudocode.
            .3 src/interpreter/parser.c.
        .2 src.
            .3 main.c.
            .3 interpreter.
                .4 bytecode.c.
                .4 bytecode.h.
                .4 error.
                    .5 error.c.
                    .5 error.h.
                .4 interpreter.c.
                .4 interpreter.h.
                .4 libraries.
                    .5 libraries.c.
                    .5 libraries.h.
                    .5 stdlib.
                        .6 stdlib.c.
                        .6 stdlib.h.
                .4 localizer.c.
                .4 localizer.h.
                .4 mappings.
                    .5 operations.c.
                    .5 operations.h.
                    .5 mapop.c.
                    .5 mapop.h.
                .4 memory.
                    .5 hashtable.c.
                    .5 hashtable.h.
                    .5 array.c.
                    .5 array.h.
                    .5 stack.c.
                    .5 stack.h.
                .4 parser.c.
                .4 parser.h.
                .4 processing.
                    .5 implementations.c.
                    .5 implementations.h.
                .4 runtime.c.
                .4 runtime.h.
                .4 tokenizer.c.
                .4 tokenizer.h.
    }
\end{flushleft}

As visible in the file tree, most of the source-code files, ending in .c,
have a corresponding header file ending in .h. This is due to the nature
of the compilation process of the C language\cte{c_compilation}, which is the host language
the \name interpreter is written in.
Each header file contains only information about functions, types or variables
available to other source or header files. The actual implementations or 
variable values are stored in the respective .c files.
% TODO:explain C compilation or link example

@@

\subsubsection{Interpreter Construction}
This section will explain the process of constructing and the inner workings of
the \name interpreter. It is therefore the most complicated and detailed section
in this paper. Examples will be made with both pseudo code and actual source
code from the project to further illustrate the decisions made during
interpreter design. File paths
used are relative to the project's root directory.

The syntax of \name is not specified in a formal programming language grammar,
but rather implemented from scratch in the files src/interpreter/tokenizer.c 
and src/interpreter/parser.c.

@@

\subsubsection{The Interpreter}
This section deals with the file src/interpreter/interpreter.c which guides
input data through the various stages of the \name interpreter in its
function 'interpret' which is called from the program's main function.
This section does not go in detail on the following stages, see their own sections.

@@

After that, variables with names are 'localized', meaning their names are
replaced by addresses where their value is stored by src/interpreter/localizer.c.
The advantage of performing this step so early is that mapping variable names
to memory is a rather resource-intensive task with time complexity of maximally 
O(log n), n being the amount of different variables.

The syntax tree could now be interpreted directly, but is instead converted
to an intermediary internal bytecode format by src/interpreter/bytecode.c.
This step is done so that code that is ran multiple times does not convey
the problem of time overhead when traversing the syntax tree, but the current
implementation does not take advantage of this.

@@

The syntax tree in the implementation of \name consists of 'stnode' type
pointers, as defined in src/interpreter/parser.h. Each such node depending
on its 'nodetype', declared in src/interpreter/tokenizer.h,
is either a value-node storing a pointer to some data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a plus-operator to signify an addition.

The difficult part of implementing a parser is precedence. Precedence and
associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. '2+3*4' would
result in 14 and is not the same as '(2+3)*4'. Therefore, each operator is
assigned a precedence, or priority, as well as an associativity, which is
either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{'20 / 10 / 2'}. Both in C as well as in \name, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{'(20 / 10) / 2'} and not as \emph{'20 / (10 / 2)'}. Both precedence 
and associativity are properties that are defined in the file mapop.gpref for
each operator.

@@

Not to be confused with machine-native bytecode, the bytecode compiler in \name
generates byte-sequences in an internal format, that are then interpreted. A
JIT compiler would instead compile the AST down to machine-native bytecode, which
could run directly on the system's hardware and would not have to be interpreted.

To generate such a sequence of bytes, \name recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack,
which is in this case used similarly to a variable-sized list.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm'\cte{shunting_yard}, traversing the tree post-order to create
a prefix notation.

@@

In src/interpreter/bytecode.c we can see, that first, the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands as such:
        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
    \item But in \name the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \expr{\emph{'- / + 6 5 * 4 2 1'}}
        This is an ideal abstraction of how data 
        is represented internally in \name.
\end{enumerate}

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available expressions are implemented in 
src/interpreter/processing/implementations.c.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name's bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name's bytecode to resemble the following expression: 
        \expr{\emph{'- 6 * 3 4'}}
        In infix notation, this would be equivalent to 
        \expr{\emph{'(4 * 3) - 6'}.}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itsself
        and notice the 6, a number, and therefore a value. The function simply returns the 6
        to the subtraction expression. This 6 will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

That's it! With all the possible expressions available, entire programs
can now be interpreted without a problem.

@@

% TODO: add examples in /examples/
\name is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. Most actual features are reliant upon this
foundation, such as operators and the standard library built into \name.

\subsection{Syntax}
The syntax of \name can be compared to the syntax of JavaScript, except
there are no keywords at all, and everything is controlled via operators,
which are non-alphanumeric characters. Such operators include the plus-sign 
for addition, or the equals-sign for variable assignment.

Following is a piece of example \name code, printing out the first ten numbers
of the fibonacci sequence, as well as a disection of how that same code is interpreted.
\code{examples/fibonacci.nts}{1}{999}
\begin{itemize}
\item The source code begins by declaring two variables \emph{a} and \emph{b}, which
    will be utilized to keep track of the data used to calculate fibonacci numbers.
    The memory allocation is done by creating a tuple of \emph{a} and \emph{b}, as suggested by 
    the comma seperating the two. Since both variables \emph{a} and \emph{b} are not yet
    known to the interpreter, they are automatically created and assigned a value
    which has no content and the \emph{undefined} typing.
\item The two variables \emph{a} and \emph{b} are then reassigned new values on line
    2 of the program. In this case it is numbers with values zero and one respectively.
    In this example, the two expressions are seperated by a space, but they can
    also be separated by newline-characters or semicolons.
\item Another variable \emph{count} is declared and defined as number in one step on line 4.
    This variable will be used to limit the amount of numbers the program outputs.
\item Yet another variable, \emph{f}, is declared and assigned a function as its
    value on line 6. In \name, normal functions do not have any arguments, 
    but they can interact with variables that were created outside their scope.
\item Lines 7 to 10 hold mathematical expressions. At first, a is assigned its current value
    plus the current value of b. Then in line 8, b is assigned the now new value of a minus the
    current value of b, which corresponds to the value of a before it was just reassigned.
    Lastly the variable \emph{count} is decremented by one.
\item Now come the interesting parts of this program.
    On lines 12 and 13, the function \emph{std.print} is referenced. This function is implemented
    in plain C code, which internally is not handled the same as regular functions in \name,
    and is wrapper to C's function \emph{printf}. The \emph{std.print} function
    takes any \name variable or value and prints it to standard output. On line 13, the "\\n"
    is replaced by a newline-character, as \name can interpret Java-like escape-sequences.
\item Another standard-library function is \emph{std.if}. It takes two or three arguments.
    If the first argument is true, then the second, else the the third argument or 
    \emph{undefined} is returned. A value is true, only if it is a number not equal to zero
    or a non-empty string. In the case of this example, \emph{std.if} returns \emph{f} if 
    count is greater than zero, and an empty function if count reaches zero. The two
    brackets following the call of the \emph{std.if} function signal yet another call to the
    returned value.
\item The final thing left to do is to get the ball roling by calling the function \emph{f} for
    the first time. As stated above, it will now print numbers and recursively call itsself,
    until the variable \emph{count} reaches a value of zero.
\end{itemize}

\subsection{Security}
Here is another example, which in contrast to the example above, focuses on a
security concept implemented in \name.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string, and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number six holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

That is why in most cases, the \emph{std.exec} function can be a risk waiting to be exploited
by malicious \name programs. To solve this problem, there is the hashtag-operator, whose purpose
it is to delete access to a variable through the specified name. It is used on line 13 of the
example to remove access to the \emph{std.exec} function, which is why the call to that function
on line 16, trying to shut down the computer, will result in an error. Unexpectedly though, the
call on line 20 to the function \emph{f} which was created as an as to \emph{std.exec} will work.

This security concept enforces the use of a top-down programming model. First, safe-to-use aliases
to dangerous functions have to be created. Then, access to all those exploitable functions is
revoked. This way, running unknown code can only access data and functions that are being provided.
This contrasts the models of languages like Python. In Python, each seperate source file can
make use of the \emph{import} keyword to include other files. Files imported this way do not
have access or knowledge about the variables that exist in the file that includes them.
In \name, each imported file would instead have access to all variables and functions that
are made available to it, instead of having to include or import them seperately.
This leads to another problem, but that is what the next section will be all about.

\subsection{Name Shadowing}
% TODO

@@

\subsection{Limitation}
\name is a programming language without a doubt. But it is not perfect in any
way. There are limitations that make it practically unusable for actual
professional projects. In this section, a few shortcomings of \name are listed
and reasoned over.

@@

\subsubsection{Memory Leaks}
Memory leaks happen, when a program reserves memory to store some data, but
never frees up the space for use for other programs, even though the data
has long been discarded and is of no further use.
In \name, memory leaks are a major problem. Even though barely noticable for small
programs it interprets, \name does not have proper setups in place to free up
all used memory that is not used any longer.

Other programming languages, notably object-oriented languages such as Java,
use Garbage-Collection, which automatically tracks unused allocated memory and
frees it up automaticadlly. While this is also the case with \name, there are
many internal flaws that have nothing to do with variables used in a \name
program, but rather internally.

A major reason for the memory leaks, but not an excuse, is the nature of the
C programming language the \name interpreter was written in. The C language and
compiler do not keep track of states in which certain blocks of memory are
forgotten about and warn the developer, instead, memory leaks happen.
More modern C alternatives such as Rust have a ownership-system in place,
in which any allocated memory has a function as an owner, and when the function
is done processing, the memory is automatically freed. Such systems are a
helpful tool for developers to create better programs with low mental overhead,
and are what programming language design should be all about.

\subsubsection{Single-Threaded}
The \name interpreter does not implement any multi-threaded, meaning parallel
running, programs as most other languages do. Instead, any \name program can
only be interpreted and run sequentially. Since most modern computer processors
are equipped with a handful of processing cores and threads, not being able to
make use of those is a major downside and in itsself a reason not to use this
language.

\subsubsection{Token input size}
In the current iteration, each token's content's size is limited to 256 bytes,
which makes it impossible to use e.g. variable names longer thatn 256 ASCII
characters. This would probably be the easyest thing to fix of here-listed 
problems of \name, but since 256 characters are long enough for 
illustration purposes, fixing this is not worth spending time on.

\subsubsection{Lack of a Development Environment}
This is certainly the greatest problem of \name, aside the countless smaller
bothering aspects.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage{listings,xcolor,setspace,hyperref,dirtree}
\usepackage{catchfile,pgffor,graphicx,float,caption}
\usepackage[margin=2.5cm]{geometry} % Set page margin to 2.5cm
\usepackage{times} % Use Times New Roman font
\usepackage[style=alphabetic,backend=biber]{biblatex}

\addbibresource{paper/bibliography.bib}

@@

% Read contents of files.txt into \files
\CatchFileDef{\files}{paper/files.txt}{\endlinechar=-1}

% Include C code from file #1 from lines #2 to #3
\newcommand{\code}[3] {
    \lstset{basicstyle=\scriptsize,keywordstyle=\color{blue},stepnumber=5,
        stringstyle=\color[rgb]{.3,.5,.1},commentstyle=\color{gray},
        morecomment=[l][\color{magenta}]{\#}}
    \lstinputlisting[language=C,numbers=left,
        frame=single,title=#1,
        firstnumber=#2,firstline=#2,lastline=#3]{#1}
}

% remove heading from lof
\makeatletter
\renewcommand\listoffigures{%
        \@starttoc{lof}%
}
\makeatother

% set caption width
\captionsetup{width=.7\linewidth}

@@

\newcommand{\paste}[1]{
    \begin{tabular}{c}
    \lstinputlisting[inputencoding=utf8,extendedchars=true,
        basicstyle=\tiny\bfseries,language={}]{#1}
    \end{tabular}
}

% the name of the programming language
\newcommand{\name}{\emph{TestScript}}
\newcommand{\pagelabel}[1]{\phantomsection\label{#1}}

@@

    \vspace*{1.5cm}
    \Huge
    \textbf{The Creation of a Programming Language}

    \vspace{1.5cm}

    % Include Mandelbrot on title page
    \begin{figure}[H]
        \centering
        \paste{paper/mandelbrot.tex}
        \caption{ASCII representation of the Mandelbrot Set, generated with \name{}}
        \label{f_mandelbrot}
    \end{figure}

    \Large
    \vspace{1.5cm}
    Matura Paper by\\ \textbf{Florian Donnelly}

@@

\normalsize

@@

@@

A programming language is itself only a description or specification, i.e. a
set of grammar rules defining how single instructions and entire programs
can be written. Source code, which is a program written in a certain language,
has to be understood by the computer hardware in order to run. It can either
be translated to machine-native bytecode with the help of a compiler, or
read line-by-line and interpreted by another program known as interpreter.
Such compilers or interpreters are programs on their own that are classified
as programming language implementation. The focus of this paper will be to
create such a programming language implementation, more specific an interpreter
of the custom-made \name{} language.

\subsection{Goals}
The focus of this project is learning about the internal design of interpreters
by creating a custom interpreter named \name{} interpreter, building upon
known strategies and algorithms used in existing programming languages.
The goal of this paper is to further understand the inner workings of
interpreters, stepping through human-readable source code and computing
desired results.

Modern programming languages often times provide additional tools aside from
the implementation. This includes dependency management tools to
easily create projects extending upon existing ones, or a web platforms for
people to share their code, all of which is not subject of this paper.
The \name{} implementation will also not include many features seen in modern
languages, as those are often created in years of time spent planning and
programming.

@@

        Self-interpreters are closely related to self-hosting compilers. These
        are in turn compilers which are written in the language they compile.
        An example of a compiler written in itself is rustc, the Rust
        compiler\cte{rustc}, which was bootstrapped by the OCaml language.
\end{itemize}
This paper will implement a mixture form of an AST compiler and bytecode
interpreter. The advantage is that an AST carries along lots of information
not needed in later stages of interpretation which can be discarded. The AST
is converted to an intermediary bytecode format which only contains
necessary information about the program. Such bytecode can then be cached,
e.g. kept in memory, meaning it only has to be converted once and can be
used many times, also leading to a performance benefit.

@@

\subsection{The Art of Simplification}
When dealing with problems in general, it is helpful to map out and think about
the way of approaching the problem first. A common approach is to distill the
problem situation down to a simpler one and extending the solution to take into
account all important aspects later.

@@

\subsection{A Project Overview}
In the following paragraphs, the project in its entirety will be briefly
presented, giving an overview over files and correlations.

\subsubsection{File Structure}
\begin{minipage}{\textwidth}

Here is a file tree overview of the project which will be referenced
several times in the following sections of this paper. In the depicted file tree
only source files are shown. No build files, executable files or files that
are of no further importance regarding \name are depicted.
\begin{flushleft}
    \scriptsize\setlength{\DTbaselineskip}{5pt}
    \dirtree{%
        .1 project root.
        .2 .gitignore.
        .2 README.md.
        .2 Makefile.
        .2 mapop.gperf.
        .2 pseudocode.
            .3 src/interpreter/parser.c.
        .2 src.
            .3 main.c.
            .3 interpreter.
                .4 bytecode.c.
                .4 bytecode.h.
                .4 error.
                    .5 error.c.
                    .5 error.h.
                .4 interpreter.c.
                .4 interpreter.h.
                .4 libraries.
                    .5 libraries.c.
                    .5 libraries.h.
                    .5 stdlib.
                        .6 stdlib.c.
                        .6 stdlib.h.
                .4 localizer.c.
                .4 localizer.h.
                .4 mappings.
                    .5 operations.c.
                    .5 operations.h.
                    .5 mapop.c.
                    .5 mapop.h.
                .4 memory.
                    .5 hashtable.c.
                    .5 hashtable.h.
                    .5 array.c.
                    .5 array.h.
                    .5 stack.c.
                    .5 stack.h.
                .4 parser.c.
                .4 parser.h.
                .4 processing.
                    .5 implementations.c.
                    .5 implementations.h.
                .4 runtime.c.
                .4 runtime.h.
                .4 tokenizer.c.
                .4 tokenizer.h.
    }
\end{flushleft}

\end{minipage}

As visible in the file tree, most of the source-code files, ending in .c,
have a corresponding header file ending in .h. This is due to the nature
of the compilation process of the C language\cte{c_compilation}, which is the host language
the \name{} interpreter is written in.
Each header file contains only information about functions, types or variables
available to other source or header files. The actual implementations or 
variable values are stored in the respective .c files.
% TODO:explain C compilation or link example

@@

\subsubsection{Flowchart}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\textwidth]{paper/Data Flow Diagram.png}
    \caption{A \name{} Data Flow Diagram}
    \label{f_diagram}
\end{figure}

The above figure is a representation of how different components of the entire
\name{} implementation interact and work together. Each next stage performs
operations on the output data of the proper. The src/interpreter/interpreter.c
file, here depicted as 'Interpreter', chains all stages together.

\subsection{Implementing \name{}}
This section will explain the process of constructing and the inner workings of
the \name{} interpreter. It is therefore the most complicated and detailed section
in this paper. Examples will be made with both pseudo code and actual source
code from the project to further illustrate the decisions made during
interpreter design. The file paths used are relative to the project's root directory.

The syntax of \name{} is not specified in a formal programming language grammar,
but rather implemented from scratch in the files src/interpreter/tokenizer.c 
and src/interpreter/parser.c.

@@

\subsubsection{The Interpreter}
This section deals with the file src/interpreter/interpreter.c which guides
input data through the various stages of the \name{} interpreter in its
function 'interpret', which is called from the program's main function.

@@

After that, variables with names are 'localized', meaning their names are
mapped to addresses in memory where their value is stored. Doing this before
bytecode compilation has a performance benefit. 
The mapping is done with a custom hashtable datastructure with average time
complexity of O(log n), n being the amount of different variables.

The syntax tree could now be interpreted directly, but as said, is instead converted
to an intermediate internal bytecode format by src/interpreter/bytecode.c.
This step is done so that code that is run multiple times does not convey
the problem of time overhead when traversing the syntax tree, but the current
implementation does not take advantage of this.

@@

The syntax tree in the implementation of \name{} consists of 'stnode' type
pointers, as defined in src/interpreter/parser.h. Each such node depending
on its 'nodetype', declared in src/interpreter/tokenizer.h,
is either a value-node storing a pointer to some data, e.g. a number or text, or a
parent-node referencing two child nodes and an operator, e.g. two numbers as
children and a plus-operator to signify an addition.

The difficult part of implementing a parser is precedence. Precedence and
associativity define the order of execution of statements. A simple example
would be to perform multiplication before addition, e.g. '2+3*4' would
result in 14 and is not the same as '(2+3)*4'. Therefore, each operator is
assigned a precedence, or priority, as well as an associativity, which is
either 'Left to Right' or 'Right to Left'. An example of associativity is
\emph{'20 / 10 / 2'}. Both in C as well as in \name{}, the division operator is
left-to-right associative, meaning the stated expression is the same as
\emph{'(20 / 10) / 2'} and not as \emph{'20 / (10 / 2)'}. Both precedence 
and associativity are properties that are defined in the file mapop.gpref for
each operator.

@@

Not to be confused with machine-native bytecode, the bytecode compiler in \name{}
generates byte-sequences in an internal format, that are then interpreted. A
JIT compiler would instead compile the AST down to machine-native bytecode, which
could run directly on the system's hardware and would not have to be interpreted.

To generate such a sequence of bytes, \name{} recursively reads through the
binary tree created by the parser and pushes found instructions onto a stack,
which is in this case used similarly to a variable-sized list.
The algorithm used is inspired by Edsger Dijkstra's 'Shunting yard 
algorithm'\cte{shunting_yard}, traversing the tree post-order to create
a prefix notation.

@@

In src/interpreter/bytecode.c we can see, that first, the left operand and then
the right operand are added onto the stack of bytecode. Therefore, when
popping from the stack and evaluating instructions, the operands will have to
be read in reverse.
Here is an example:
\begin{enumerate}
    \item Assume the following mathematical expression in infix notation:
        \expr{\emph{'1 - 2 * 4 / (5 + 6)'}}
    \item In common prefix notation, also called Polish notation, the
        operators would now precede the operands as such:
        \expr{\emph{'- 1 (/ (* 2 4) (+ 5 6))'}}
    \item But in \name{} the operands are pushed onto the stack is reverse order.
        This would look like the following:
        \expr{\emph{'- (/ (+ 6 5) (* 4 2)) 1'}}
    \item The brackets are not needed, because for every operator, the amount
        of operands is known. This is the case as e.g. the subtraction operator
        and the negative operator, which are both a minus sign, are distinguished
        beforehand. This means the above statement would look like this:
        \expr{\emph{'- / + 6 5 * 4 2 1'}}
        This is an ideal abstraction of how data 
        is represented internally in \name{}.
\end{enumerate}

\subsubsection{The Runtime}
This is the last stage of interpreting program code as implemented in \name{}.
The runtime steps through the intermediary bytecode created by the bytecode
compiler and evaluates expressions by plugging in the provided values.
All available expressions are implemented in 
src/interpreter/processing/implementations.c.

The evaluation of expressions also works in a recursive manner. Here is an 
example of \name{}'s bytecode notation and how it would be evaluated at runtime:
\begin{itemize}
    \item Assume the input in \name{}'s bytecode to resemble the following expression: 
        \expr{\emph{'- 6 * 3 4'}}
        In infix notation, this would be equivalent to 
        \expr{\emph{'(4 * 3) - 6'}.}
    \item The runtime will read through the bytecode left-to-right. First it will
        notice the subtraction-operator and accordingly assume that two
        operands must follow.
    \item The processing function will call itsself
        and notice the 6, a number, and therefore a value. The function simply returns the 6
        to the subtraction expression. This 6 will be the right-hand operator
        of the subtraction.
    \item The subtraction will then call the processing function again to find
        the left operand. As the processing function reads a multiplication operator,
        it now recursively reads through that expression's right and then left
        operator, which would be 3 and for respectively and compute and return
        the result of the multiplication, which is 12.
    \item Now the subtraction knows its right- and left-hand operators, 6 and 12,
        and again, calculate and return the result, which is 6.
\end{itemize}

That is it. With all the expressions available, entire programs
can now be interpreted.

@@

% TODO: add examples in /examples/
\name{} is built upon the implementation of a recursive Pratt
parser and a basic tokenizer. Most actual features are reliant upon this
foundation, such as operators and the standard library built into \name{}.

\subsection{Syntax}
The syntax of \name{} can be compared to the syntax of JavaScript, except
there are no keywords at all, and everything is controlled via operators,
which are non-alphanumeric characters. Such operators include the plus-sign 
for addition, or the equals-sign for variable assignment.

Following is a piece of example \name{} code, printing out the first ten numbers
of the fibonacci sequence, as well as a disection of how that same code is interpreted.
\code{examples/fibonacci.nts}{1}{999}
\begin{itemize}
\item The source code begins by declaring two variables \emph{a} and \emph{b}, which
    will be utilized to keep track of the data used to calculate fibonacci numbers.
    The memory allocation is done by creating a tuple of \emph{a} and \emph{b}, as suggested by 
    the comma seperating the two. Since both variables \emph{a} and \emph{b} are not yet
    known to the interpreter, they are automatically created and assigned a value
    which has no content and the \emph{undefined} typing.
\item The two variables \emph{a} and \emph{b} are then reassigned new values on line
    2 of the program. In this case it is numbers with values zero and one respectively.
    In this example, the two expressions are seperated by a space, but they can
    also be separated by newline-characters or semicolons.
\item Another variable \emph{count} is declared and defined as number in one step on line 4.
    This variable will be used to limit the amount of numbers the program outputs.
\item Yet another variable, \emph{f}, is declared and assigned a function as its
    value on line 6. In \name{}, normal functions do not have any arguments, 
    but they can interact with variables that were created outside their scope.
\item Lines 7 to 10 hold mathematical expressions. At first, a is assigned its current value
    plus the current value of b. Then in line 8, b is assigned the now new value of a minus the
    current value of b, which corresponds to the value of a before it was just reassigned.
    Lastly the variable \emph{count} is decremented by one.
\item Now come the interesting parts of this program.
    On lines 12 and 13, the function \emph{std.print} is referenced. This function is implemented
    in plain C code, which internally is not handled the same as regular functions in \name{},
    and is wrapper to C's function \emph{printf}. The \emph{std.print} function
    takes any \name{} variable or value and prints it to standard output. On line 13, the "\\n"
    is replaced by a newline-character, as \name{} can interpret Java-like escape-sequences.
\item Another standard-library function is \emph{std.if}. It takes two or three arguments.
    If the first argument is true, then the second, else the the third argument or 
    \emph{undefined} is returned. A value is true, only if it is a number not equal to zero
    or a non-empty string. In the case of this example, \emph{std.if} returns \emph{f} if 
    count is greater than zero, and an empty function if count reaches zero. The two
    brackets following the call of the \emph{std.if} function signal yet another call to the
    returned value.
\item The final thing left to do is to get the ball roling by calling the function \emph{f} for
    the first time. As stated above, it will now print numbers and recursively call itsself,
    until the variable \emph{count} reaches a value of zero.
\end{itemize}

\subsection{Name Shadowing}
Here is another example, which in contrast to the example above, focuses on a
security concept implemented in \name{}.
\code{examples/exec.nts}{1}{999}
The code begins by creating the variable \emph{var}, which holds a string, and
\emph{f}, which calls the function \emph{std.exec} with var as argument.
Line number six holds a comment, indicated by the two slashes at the start of the line.
Comments are ignored by the \name{} interpreter and their purpose is to contain notes
and explanations inside the source code itself. Another type of commends are block-comments
which span across multiple lines, such as on lines 9 to 12.
On line seven, the \emph{std.exec} function is called for the first time. It runs the
command it is given as argument in the user's shell, which is most commonly \emph{cmd} on
Microsoft Windows and \emph{bash} on most Linux distributions. In this example, \emph{std.exec}
launches the Firefox browser, but it is capable of all kinds of things that can otherwise be
done in the shell, such as delete files or shut down the computer.

That is why in most cases, the \emph{std.exec} function can be a risk waiting to be exploited
by malicious \name{} programs. To solve this problem, there is the hashtag-operator, whose purpose
it is to delete access to a variable through the specified name. It is used on line 13 of the
example to remove access to the \emph{std.exec} function, which is why the call to that function
on line 16, trying to shut down the computer, will result in an error. Unexpectedly though, the
call on line 20 to the function \emph{f} which was created as an as to \emph{std.exec} will work.

This security concept enforces the use of a top-down programming model. First, safe-to-use aliases
to dangerous functions have to be created. Then, access to all those exploitable functions is
revoked. This way, running unknown code can only access data and functions that are being provided.
This contrasts the models of languages like Python. In Python, each seperate source file can
make use of the \emph{import} keyword to include other files. Files imported this way do not
have access or knowledge about the variables that exist in the file that includes them.
In \name{}, each imported file would instead have access to all variables and functions that
are made available to it, instead of having to include or import them seperately.
This leads to another problem, but that is what the next section will be all about.

@@

\subsection{Limitations}
\name{} is a programming language without a doubt. But it is not perfect in any
way. There are limitations that make it practically unusable for actual
professional projects. In this section, a few shortcomings of \name{} are listed
and reasoned over.

@@

\subsubsection{Memory Leaks}
Memory leaks happen, when a program reserves memory to store some data, but
never frees up the space for use for other programs, even though the data
has long been discarded and is of no further use.
In \name{}, memory leaks are a major problem. Even though barely noticable for small
programs it interprets, \name{} does not have proper setups in place to free up
all used memory that is not used any longer.

Other programming languages, notably object-oriented languages such as Java,
use Garbage-Collection, which automatically tracks unused allocated memory and
frees it up automaticadlly. While this is also the case with \name{}, there are
many internal flaws that have nothing to do with variables used in a \name{}
program, but rather internally.

A major reason for the memory leaks, but not an excuse, is the nature of the
C programming language the \name{} interpreter was written in. The C language and
compiler do not keep track of states in which certain blocks of memory are
forgotten about and warn the developer, instead, memory leaks happen.
More modern C alternatives such as Rust have a ownership-system in place,
in which any allocated memory has a function as an owner, and when the function
is done processing, the memory is automatically freed. Such systems are a
helpful tool for developers to create better programs with low mental overhead,
and are what programming language design should be all about.

\subsubsection{Single-Threaded}
The \name{} interpreter does not implement any multi-threaded, meaning parallel
running, programs as most other languages do. Instead, any \name{} program can
only be interpreted and run sequentially. Since most modern computer processors
are equipped with a handful of processing cores and threads, not being able to
make use of those is a major downside and in itsself a reason not to use this
language.

\subsubsection{Lack of a Development Environment}
This is certainly the greatest problem of \name{}, aside the countless smaller
bothering aspects.

Most modern programming languages do not only consist
of a language specification and a compiler or interpreter, but of many more
tools surrounding them.
Such tools can include a package-manager, with which people around the world
can share code. Examples would be npm\cte{npm} or cargo\cte{cargo}, 
the 'Node Package Manager' for JavaScript applications and and Rust's package 
manager respectively.

\subsection{Interesting Ideas untouched}
Creating the ideal programming language takes a lot of planning and sketching
out. This section of the paper gives insight on interesting ideas or concepts
that are great to have and are used in other programming languages, but not 
could not be implemented into \name{}.
